### 2009-06-16 paper.tex
Moved algorithms into main text

In our tests, incorporation of basepair structure was the most important factor for accurate inference of alignment;
appropriate use of posterior-decoding was the next most important factor;
and fine details of the model were least important.
Since posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible,
with transducer composition reserved for exact ML (or MCMC) inference on phylogenies.
We discuss possible ways to make these phylogenetic inferences tractable.

@@

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in memory and
time (respectively) for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
(such as a curated multiple alignment)
about the structures and/or evolutionary histories of the sequences of
interest.  We can use this knowledge as a constraint to reduce the accessible volume, and hence the storage requirements, of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelope'' concept.

@@

We use two orderings for sequences in the fold envelopes.  An inside-outside ordering is used for the iteration 
in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
contains all previous subsequences in the fold envelope.  More precisely, subsequences in $\foldenv^{(X)}$ are sorted  
in the same order as coordinate pairs $(i,j)$ are generated by the
iteration $\{\,\mathrm{for}\, i = L^{(X)} \,\mathrm{to}\, 0 \,\,\{
\mathrm{for}\, j = i \,\mathrm{to}\, L^{(X)} \}\,\}$.

The Outside algorithm uses an outside-inside ordering, which is the exact reverse of the inside-outside ordering described above.
Subsequences in $\foldenv^{(X)}$ are sorted 
in the same order as coordinate pairs $(i,j)$ are generated by the 
iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L^{(X)} \,\,\{ \mathrm{for}\, j = L^{(X)}\, \mathrm{to}\, i \}\,\}$.

@@

@@

@@

@@

@@

\clearpage
\section*{Algorithms}

  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

  \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside-outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

  \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{outside-inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{b}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \caption{\alglabel{outside}
    The constrained Outside algorithm for three sequences $X, Y, Z$.
    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
    with $\Semit$ states first, then $\Snull$ states in topological order.
  }
\end{algorithm}

>>>>

In our tests, incorporation of basepair structure was the most important factor for accurate inference of alignment;
appropriate use of posterior-decoding was the next most important factor;
and fine details of the model were least important.
Posterior-decoding heuristics can be substantially faster than exact phylogenetic inference,
so this motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics where possible.
We discuss the use of transducer composition for ML (or MCMC) inference on phylogenies,
including possible ways to make these phylogenetic inferences tractable.

@@

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in memory and
time (respectively) for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
(such as a curated multiple alignment)
about the structures and/or evolutionary histories of the sequences of
interest.  We can use this knowledge as a constraint to reduce the accessible volume, and hence the storage requirements, of the DP matrix \cite{Holmes2005}.
The Inside, Outside, and CYK+traceback algorithms for a three-taxon star phylogeny
can be constrained using the ``fold envelope'' concept, which will now be described.

@@

An inside$\to$outside ordering is used for the iteration 
in the Inside algorithm: Subsequences are ordered such that each successive subsequence 
contains all previous subsequences in the fold envelope.  More precisely, subsequences in $\foldenv^{(X)}$ are sorted  
in the same order as coordinate pairs $(i,j)$ are generated by the
iteration $\{\,\mathrm{for}\, i = L^{(X)} \,\mathrm{to}\, 0 \,\,\{
\mathrm{for}\, j = i \,\mathrm{to}\, L^{(X)} \}\,\}$.

The Outside algorithm uses the exact reverse of the inside$\to$outside ordering described above; we call this the outside$\to$inside ordering.
Subsequences in $\foldenv^{(X)}$ are sorted 
in the same order as coordinate pairs $(i,j)$ are generated by the 
iteration $\{\,\mathrm{for}\, i = 0 \,\mathrm{to}\, L^{(X)} \,\,\{ \mathrm{for}\, j = L^{(X)}\, \mathrm{to}\, i \}\,\}$.

@@

Using the fold envelope formalism, the main iteration over cells in the Inside and CYK matrices can be expressed as three nested loops:
one for each sequence, traversing the fold envelope subsequences in inside$\to$outside order.
Conversely, the main iteration of the Outside algorithm consists of three nested outside$\to$inside loops.

@@

  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ $\leftarrow$ \calcTransEmitProb$\left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\alpha_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\alpha_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

@@

  \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{inside$\to$outside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{a}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_L,n^{(X)}_R \right) \in b_{in}\left( n^{(X)} \right)$} {
            \ForEach{$\left( n^{(Y)}_L,n^{(Y)}_R \right) \in b_{in}\left( n^{(Y)} \right)$} {
              \ForEach{$\left( n^{(Z)}_L,n^{(Z)}_R \right) \in b_{in}\left( n^{(Z)} \right)$} {
                \bifurcProb += \calcLBifurcProb($\bvec{a}; \cdot$)\;
                \bifurcProb += \calcRBifurcProb($\bvec{a}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ \\
          $\qquad \leftarrow \max \left( \calcTransEmitProb \left( \bvec{a}; n^{(X)}, n^{(Y)}, n^{(Z)} \right), \bifurcProb \right)$ \;
          store $\gamma_{\bvec{a}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \KwRet{$\gamma_{\bvec{a}}\left( n^{(X)}[0,L^{(X)}],n^{(Y)}[0,L^{(Y)}],n^{(Z)}[0,L^{(Z)}] \right)$}\;

@@

  \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(X)} \in \foldenv^{(X)}$} {
    \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Y)} \in \foldenv^{(Y)}$} {
      \ForEach(\tcc*[f]{outside$\to$inside sorted}){$n^{(Z)} \in \foldenv^{(Z)}$} {
        \BlankLine
        \ForEach{state $\bvec{b}$}{
          \BlankLine
          \bifurcProb $\leftarrow 0$\;
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_L \right) \in b_{out,L}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_L \right) \in b_{out,L}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_L \right) \in b_{out,L}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcLBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \ForEach{$\left( n^{(X)}_O,n^{(X)}_R \right) \in b_{out,R}\left( n^{(X)} \right)$}{
            \ForEach{$\left( n^{(Y)}_O,n^{(Y)}_R \right) \in b_{out,R}\left( n^{(Y)} \right)$}{
              \ForEach{$\left( n^{(Z)}_O,n^{(Z)}_R \right) \in b_{out,R}\left( n^{(Z)} \right)$}{
                \bifurcProb += \calcRBifurcProb($\bvec{b}; \cdot$)\;
              }
            }
          }
          \BlankLine
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right) \leftarrow$ \calcTransEmitProb$\left( \bvec{b}; n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
          $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$ += \bifurcProb\;
          store $\beta_{\bvec{b}} \left( n^{(X)}, n^{(Y)}, n^{(Z)} \right)$\;
        }
        \BlankLine
      }
    }
  }
  \caption{\alglabel{outside}
    The constrained Outside algorithm for three sequences $X, Y, Z$.
    Ensemble states $\bvec{a}$ in the iteration over states are sorted in Outside fill order
    with $\Semit$ states first, then $\Snull$ states in topological order.
  }
\end{algorithm}

@@
